---
title: "RPl-MODISTools"
output: html_document
---
This study is a replication of the MODISTools vignette (located [here](https://cran.r-project.org/web/packages/MODISTools/vignettes/modistools-vignette.html)), which is an analysis of LAI (leaf area index) near the Bay of Arcachon, France. Additional code about using `mt_batch_subset` was obtained from Christian John's 2019 tutorial, found [here](https://jepsonnomad.github.io/tutorials/MODISTools_Intro.html). 

The full MODISTools documentation can be found [here](https://cran.r-project.org/web/packages/MODISTools/MODISTools.pdf).

## Load packages
```{r load packages}
# list of required packages
packages = c("here", "MODISTools", "tidyverse", "raster", "dplyr", "stars", "USAboundaries", "ggrepel")

# load and install required packages
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# save the R processing environment to r_environment.txt
writeLines(capture.output(sessionInfo()),here("procedure","environment","r_environment.txt"))

```

## Finding data
There are many products (including bands and dates) availablefor access using MODISTools. To view all the available products from the MODIS sensor, with description, return interval, and resolution, run the following code:
```{r view available products from MODIS + description}
products <- mt_products()
head(products)
```
To view available bands for a given product:
```{r view available bands for a product}
bands <- mt_bands(product = "MOD15A2H")
head(bands)
```
To view available dates for a given product at a particular location:
```{r view available dates for a product}
dates <- mt_dates(product = "MCD12Q1", lat = 32, lon = -117)
head(dates)
```

## Create study area figure

```{r plot study area within California}
ROI = data.frame("site_name" = c("Study area"),
                  "lat" = c(32.674),
                  "lon" = c(-116.8054))

# Access CA boundary shape
CA = USAboundaries::us_states(resolution = "low",
                              states = "CA")
# Plot
ggplot() +
  geom_sf(data = CA) +
  geom_point(data = ROI,
             aes(x = lon, y = lat)) +
  geom_text_repel(data = ROI,
                  aes(x = lon,
                      y = lat,
                      label = site_name))

ggsave(here("results", "figures", "study_area.png"))
```

## Download LAI timeseries
Use `mt_batch_subset` to download timeseries over longer timescales. 
```{r download MODIS LAI data 2003-2021}
dlStart = Sys.time()
LAI_full = mt_batch_subset(df = ROI,
                           product = "MOD15A2H",
                           band = "Lai_500m",
                           internal = TRUE,
                           start = "2003-01-01",
                           end = "2021-12-31")
dlEnd = Sys.time()
dlEnd - dlStart
```
 
```{r plot LAI 2003-2021}
ggplot(LAI_full,
       aes(x = as.POSIXct(calendar_date),
           y = value)) + 
  geom_point() +
  geom_line() +
  geom_smooth() +
  ylab("LAI") +
  xlab("Date") +
  theme_classic() +
  scale_color_brewer(palette = "Dark2")

ggsave(here("results", "figures", "lai_timeseries.png"))
```
# 2003 and 2019 comparison
## 2003
### Download data 
Use `mt_subset()` to download timeseries. 
```{r download MODIS land cover and LAI data '03}
# leaf area index
fire_lai_03 <- mt_subset(product = "MOD15A2H",
                      lat = 32.674,
                      lon =  -116.8054,
                      band = "Lai_500m",
                      start = "2003-04-01",
                      end = "2003-04-30",
                      km_lr = 20,
                      km_ab = 20,
                      site_name = "san diego",
                      internal = TRUE,
                      progress = FALSE)
# land cover
fire_lc_03 <- mt_subset(product = "MCD12Q1",
                     lat = 32.674,
                     lon =  -116.8054,
                     band = "LC_Type1",
                     start = "2003-01-01",
                     end = "2003-12-31",
                     km_lr = 20,
                     km_ab = 20,
                     site_name = "san_diego",
                     internal = TRUE,
                     progress = FALSE)
```

```{r download MODIS land cover and LAI data '19}
# leaf area index
fire_lai_19 <- mt_subset(product = "MOD15A2H",
                      lat = 32.674,
                      lon =  -116.8054,
                      band = "Lai_500m",
                      start = "2019-04-01",
                      end = "2019-04-30",
                      km_lr = 20,
                      km_ab = 20,
                      site_name = "san diego",
                      internal = TRUE,
                      progress = FALSE)
# land cover
fire_lc_19 <- mt_subset(product = "MCD12Q1",
                     lat = 32.674,
                     lon =  -116.8054,
                     band = "LC_Type1",
                     start = "2019-01-01",
                     end = "2019-12-31",
                     km_lr = 20,
                     km_ab = 20,
                     site_name = "san_diego",
                     internal = TRUE,
                     progress = FALSE)
```
### Process data
```{r merge land cover and leaf area index}
# merge land cover and lai data
fire_03 <- fire_lc_03 %>%
  rename("lc" = "value") %>%
  dplyr::select("lc","pixel") %>%
  right_join(fire_lai_03, by = "pixel")
```

```{r determine which classes are present}
fire_03 %>%
  group_by(lc) %>% 
  summarise(count = sum(lc))
```
6 = closed shrublands
7 = open shrublands
9 = savannas
10 = grasslands
11 = permanent wetlands
12 = croplands
13 = urban and builtup lands
16 = barren
17 = water bodies

"Then, filter out all non valid values (> 100), only select evergreen and deciduous land cover classes (1 and 5, or, ENF and DBF respectively), convert them to more readable labels, and across these land cover classes take the **median per acquisition date**."

```{r filter non-valid values + rename classes + take median}
# create a plot of the data - accounting for the multiplier (scale) component
fire_03 <- fire_03 %>%
  filter(value <= 100,
         lc %in% c("6", "7", "9", "10", "11", "12", "13", "16")) %>% # retain everything but fill values
  mutate(lc = case_when(
    lc == 6 ~ "closedshrub",
    lc == 7 ~ "openshrub",
    lc == 9 ~ "savanna",
    lc == 10 ~ "grassland",
    lc == 11 ~ "wetland",
    lc == 12 ~ "cropland",
    lc == 13 ~ "urbanland",
    lc == 16 ~ "barren")) %>% 
  group_by(lc, calendar_date) %>% # group by lc and date
  summarize(doy = as.numeric(format(as.Date(calendar_date)[1],"%j")),
            lai_mean = median(value * as.double(scale)))
```

"Finally, the plot will show you the seasonal time series of LAI for both land cover classes (ENF and DBF). Note the difference in timing and amplitude between both these forest types, where the evergreen (ENF) pixels show lower LAI values and a more gradual seasonal pattern compared to the deciduous trees."

```{r plot LAI by date and per land cover class}
# plot LAI by date and per land cover class
ggplot(fire_03, aes(x = doy, y = lai_mean)) +
  geom_point() +
  geom_smooth(span = 0.3, method = "loess") +
  labs(x = "day of year (DOY)",
       y = "leaf area index (LAI)") +
  theme_minimal() +
  facet_wrap(~ lc, scales = "free")
```

```{r sinusoidal to lat-long}
# convert the coordinates from sinusoidal to lat-long
lat_lon_03 <- sin_to_ll(fire_lc_03$xllcorner, fire_lc_03$yllcorner)

# bind with the original dataframe
subset_03 <- cbind(fire_lc_03, lat_lon_03)

head(subset_03)
```

```{r calculate bounding box of extracted data}
# convert to bounding box
bb_03 <- apply(fire_lc_03, 1, function(x){
  mt_bbox(xllcorner = x['xllcorner'],
          yllcorner = x['yllcorner'],
           cellsize = x['cellsize'],
           nrows = x['nrows'],
           ncols = x['ncols'])
})

# plot one bounding box
plot(bb_03[[1]])

# add the location of the queried coordinate within the polygon
points(fire_lc_03$longitude[1],
       fire_lc_03$latitude[1],
       pch = 20,
       col = "red")
```

### Convert to gridded raster data
```{r convert to raster & plot}
# convert to raster, when reproject is TRUE
# the data is reprojected to lat / lon if FALSE
# the data is shown in its original sinuidal projection
LC_r_03 <- mt_to_raster(df = fire_lc_03, reproject = FALSE)
LAI_r_03 <- mt_to_raster(df = fire_lai_03, reproject = FALSE)


# plot the raster data as a map
plot(LC_r_03)
plot(LAI_r_03)
```


## 2019 (same process as above)
### Process data
```{r merge land cover and leaf area index}
# merge land cover and lai data
fire_19 <- fire_lc_03 %>%
  rename("lc" = "value") %>%
  dplyr::select("lc","pixel") %>%
  right_join(fire_lai_03, by = "pixel")
```

```{r determine which classes are present}
fire_19 %>%
  group_by(lc) %>% 
  summarise(count = sum(lc))
```
6 = closed shrublands
7 = open shrublands
8 = woody savanna
9 = savannas
10 = grasslands
11 = permanent wetlands
12 = croplands
13 = urban and builtup lands
16 = barren
17 = water bodies

"Then, filter out all non valid values (> 100), only select evergreen and deciduous land cover classes (1 and 5, or, ENF and DBF respectively), convert them to more readable labels, and across these land cover classes take the **median per acquisition date**."

```{r filter non-valid values + rename classes + take median}
# create a plot of the data - accounting for the multiplier (scale) component
fire_19 <- fire_19 %>%
  filter(value <= 100,
         lc %in% c("6", "7", "8", "9", "10", "11", "12", "13", "16")) %>% # retain everything but fill values
  mutate(lc = case_when(
    lc == 6 ~ "closedshrub",
    lc == 7 ~ "openshrub",
    lc == 8 ~ "woodysavanna",
    lc == 9 ~ "savanna",
    lc == 10 ~ "grassland",
    lc == 11 ~ "wetland",
    lc == 12 ~ "cropland",
    lc == 13 ~ "urbanland",
    lc == 16 ~ "barren")) %>% 
  group_by(lc, calendar_date) %>% # group by lc and date
  summarize(doy = as.numeric(format(as.Date(calendar_date)[1],"%j")),
            lai_mean = median(value * as.double(scale)))
```

"Finally, the plot will show you the seasonal time series of LAI for both land cover classes (ENF and DBF). Note the difference in timing and amplitude between both these forest types, where the evergreen (ENF) pixels show lower LAI values and a more gradual seasonal pattern compared to the deciduous trees."

Use these plots to explain why to use April as peak vegetation time!
```{r plot LAI by date and per land cover class}
# plot LAI by date and per land cover class
ggplot(fire_19, aes(x = doy, y = lai_mean)) +
  geom_point() +
  geom_smooth(span = 0.3, method = "loess") +
  labs(x = "day of year (DOY)",
       y = "leaf area index (LAI)") +
  theme_minimal() +
  facet_wrap(~ lc, scales = "free")
```

```{r sinusoidal to lat-long}
# convert the coordinates from sinusoidal to lat-long
lat_lon_19 <- sin_to_ll(fire_lc_19$xllcorner, fire_lc_19$yllcorner)

# bind with the original dataframe
subset_19 <- cbind(fire_lc_19, lat_lon_19)

head(subset_19)
```

```{r calculate bounding box of extracted data}
# convert to bounding box
bb_19 <- apply(fire_lc_19, 1, function(x){
  mt_bbox(xllcorner = x['xllcorner'],
          yllcorner = x['yllcorner'],
           cellsize = x['cellsize'],
           nrows = x['nrows'],
           ncols = x['ncols'])
})

# plot one bounding box
plot(bb_19[[1]])

# add the location of the queried coordinate within the polygon
points(fire_lc_19$longitude[1],
       fire_lc_19$latitude[1],
       pch = 20,
       col = "red")
```

### Convert to gridded raster data
```{r convert to raster & plot}
# convert to raster, when reproject is TRUE
# the data is reprojected to lat / lon if FALSE
# the data is shown in its original sinuidal projection
LC_r_19 <- mt_to_raster(df = fire_lc_19, reproject = FALSE)
LAI_r_19 <- mt_to_raster(df = fire_lai_19, reproject = FALSE)


# plot the raster data as a map
plot(LC_r_19)
plot(LAI_r_19)
```

## Save both rasters for both years
```{r write rasters to .tif file}
LC_r_03_stars <- st_as_stars(LC_r_03)
LAI_r_03_stars <- st_as_stars(LAI_r_03)

LC_r_19_stars <- st_as_stars(LC_r_19)
LAI_r_19_stars <- st_as_stars(LAI_r_19)

write_stars(LC_r_03_stars, here("data", "derived", "private", "LC_r_03.tif"))
write_stars(LAI_r_03_stars, here("data", "derived", "private", "LAI_r_03.tif"))
write_stars(LC_r_19_stars, here("data", "derived", "private", "LC_r_19.tif"))
write_stars(LAI_r_19_stars, here("data", "derived", "private", "LAI_r_19.tif"))
```


## Raster calculator
```{r calculate difference in LAI between 2019 and 2003}
test19 <- abs(max(LAI_r_19[[1]], na.rm = TRUE))
test03 <- abs(max(LAI_r_03[[1]], na.rm = TRUE))

diff <- test19 - test03

plot(test19)
plot(test03)
plot(diff)
```

## Plot LC composition over time
```{r join two lc datasets to see changes}
lc_combined <- fire_lc_19 %>%
  rename("lc19" = "value") %>%
  dplyr::select("lc19","pixel") %>%
  right_join(fire_lc_03, by = "pixel") %>% 
  rename("lc03"= "value")

# preview pixels with changed lc values
lc_combined %>% 
  dplyr::select(lc19, lc03) %>% 
  filter(lc19 != lc03)
```


```{r combine two datasets for plotting}
fire_lc_19_plot <- fire_lc_19 %>% 
  mutate(year = "2019")

fire_lc_03_plot <- fire_lc_03 %>% 
  mutate(year = "2003")

fire_toplot <- rbind(fire_lc_19_plot, fire_lc_03_plot)

fire_toplot <- fire_toplot %>% 
  mutate(lc_name = case_when(
    value == 6 ~ "closed shrub",
    value == 7 ~ "open shrub",
    value == 8 ~ "woody savanna",
    value == 9 ~ "savanna",
    value == 10 ~ "grassland",
    value == 11 ~ "wetland",
    value == 12 ~ "cropland",
    value == 13 ~ "urban land",
    value == 16 ~ "barren")) %>% 
  group_by(lc_name, year) %>% 
  summarise(num = n())
```

```{r plot 2003 vs. 2019 LC}
fire_toplot %>% 
  ggplot(aes(x = num, y = year, fill = lc_name)) + 
  geom_bar(stat = "identity") + 
  ylab("Year") + 
  xlab("Count") + 
  scale_fill_manual("Land Cover Class", values = c("closed shrub" = "#2C5530", "grassland" = "#A5BE00", "open shrub" = "#DCED31", "savanna" = "#FFC857", "urban land" = "#4A6FA5")) + 
  theme_classic()

ggsave(here("results", "figures", "lc_comparison.png"))

```