---
title: "RPl-MODISTools"
output: html_document
---
This study is a replication of the MODISTools vignette, located [here](https://cran.r-project.org/web/packages/MODISTools/vignettes/modistools-vignette.html), which is an analysis of LAI (leaf area index) near the Bay of Arcachon, France.

## Load packages
```{r load packages}
# list of required packages
packages = c("here", "MODISTools", "tidyverse", "raster", "dplyr")

# load and install required packages
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# save the R processing environment to r_environment.txt
writeLines(capture.output(sessionInfo()),here("procedure","environment","r_environment.txt"))

```

## Finding data

There are many products (including bands and dates) availablefor access using MODISTools. To view all the available products from the MODIS sensor, with description, return interval, and resolution, run the following code:
```{r}
products <- mt_products()
head(products)
```
To view available bands for a given product:
```{r}
bands <- mt_bands(product = "MCD15A3H")
head(bands)
```
To view available dates for a given product at a particular location:
```{r}
dates <- mt_dates(product = "MCD15A3H", lat = 32, lon = -117)
head(dates)
```

## Download timeseries
Use `mt_subset()` to download timeseries. 
Output is saved to variables called "subset" and "LC" in the R workspace  (as defined by the parameter internal = TRUE, when set to FALSE the data is written to file).
**Emma, make it save to file**
This query may take a while.
```{r}
# download the MODIS land cover (IGBP) and NDVI data
# for a region around the French city and basin of Arcachon

# leaf area index
fire_lai <- mt_subset(product = "MOD15A2H",
                    lat = 32.7157,
                    lon =  -117.1611,
                    band = "Lai_500m",
                    start = "2002-01-01",
                    end = "2002-12-30",
                    km_lr = 20,
                    km_ab = 20,
                    site_name = "san diego",
                    internal = TRUE,
                    progress = FALSE)
# land cover
fire_lc <- mt_subset(product = "MCD12Q1",
  lat = 32.7157,
  lon =  -117.1611,
  band = "LC_Type1",
  start = "2002-01-01",
  end = "2002-12-30",
  km_lr = 20,
  km_ab = 20,
  site_name = "san_diego",
  internal = TRUE,
  progress = FALSE)
```

```{r preview LAI}
head(arcachon_lai)
```

```{r preview LC}
head(arcachon_lc)
```

```{r}
lc_uniq <- unique(arcachon_lc[c("pixel")])
nrow(lc_uniq)
```
```{r}
lai_uniq <- unique(arcachon_lai[c("pixel")])
nrow
nrow(lai_uniq)
```




```{r}
# merge land cover and lai data
fire <- fire_lc %>%
  rename("lc" = "value") %>%
  dplyr::select("lc","pixel") %>%
  right_join(fire_lai, by = "pixel")
```

```{r}
arcachon %>%
  group_by(lc) %>% 
  summarise(count = sum(lc))
```

```{r}
fire %>%
  group_by(lc) %>% 
  summarise(count = sum(lc))
```
6 = closed shrublands
7 = open shrublands
9 = savannas
10 = grasslands
11 = permanent wetlands
13 = urban and builtup lands
16 = barren
17 = water bodies



"Then, filter out all non valid values (> 100), only select evergreen and deciduous land cover classes (1 and 5, or, ENF and DBF respectivelly), convert them to more readable labels, and across these land cover classes take the median per acquisition date."

```{r}
# create a plot of the data - accounting for the multiplier (scale) component
fire <- fire %>%
  filter(value <= 100,
         lc %in% c("6", "7", "9", "10", "11", "13", "16")) %>% # retain everything but fill values
  mutate(lc = case_when(
    lc == 6 ~ "closedshrub",
    lc == 7 ~ "openshrub",
    lc == 9 ~ "savanna",
    lc == 11 ~ "grassland",
    lc == 13 ~ "urbanlands",
    lc == 16 ~ "barren")) %>% 
  group_by(lc, calendar_date) %>% # group by lc and date
  summarize(doy = as.numeric(format(as.Date(calendar_date)[1],"%j")),
            lai_mean = median(value * as.double(scale)))
```
"Finally, the plot will show you the seasonal time series of LAI for both land cover classes (ENF and DBF). Note the difference in timing and amplitude between both these forest types, where the evergreen (ENF) pixels show lower LAI values and a more gradual seasonal pattern compared to the deciduous trees."

```{r}
# plot LAI by date and per land cover class
ggplot(fire, aes(x = doy, y = lai_mean)) +
  geom_point() +
  geom_smooth(span = 0.3, method = "loess") +
  labs(x = "day of year (DOY)",
       y = "leaf area index (LAI)") +
  theme_minimal() +
  facet_wrap(~ lc)
```
```{r}
# convert the coordinates from sinusoidal to lat-long
lat_lon <- sin_to_ll(fire_lc$xllcorner, fire_lc$yllcorner)

# bind with the original dataframe
subset <- cbind(fire_lc, lat_lon)

head(subset)

```

```{r calculate bounding box of extracted data}
# convert to bounding box
bb <- apply(fire_lc, 1, function(x){
  mt_bbox(xllcorner = x['xllcorner'],
          yllcorner = x['yllcorner'],
           cellsize = x['cellsize'],
           nrows = x['nrows'],
           ncols = x['ncols'])
})

# plot one bounding box
plot(bb[[1]])

# add the location of the queried coordinate within the polygon
points(fire_lc$longitude[1],
       fire_lc$latitude[1],
       pch = 20,
       col = "red")
```

## Convert to gridded raster data
```{r}
# convert to raster, when reproject is TRUE
# the data is reprojected to lat / lon if FALSE
# the data is shown in its original sinuidal projection
LC_r <- mt_to_raster(df = fire_lc, reproject = TRUE)

# plot the raster data as a map
plot(LC_r)
```

